---
title: "EPPS Math Coding Camp"
subtitle: "Regression Diagnostics and Model Evaluation"
author: "Dohyo Jeong"
date: "August 13, 2024"
format: 
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    code-tools: true
    highlight-style: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Session 1: Regression Diagnostics and Model Evaluation

## 1.1 Introduction and Data Overview

### Dataset: Boston from the MASS Package

The Boston dataset from the MASS package includes housing data from the Boston area, with variables such as crime rate, tax rates, and median home values. <br>

**Key variables:**<br>
- **crim**: Crime rate per capita by town.<br>
- **zn**: Proportion of residential land zoned for large lots.<br>
- **indus**: Proportion of non-retail business acres per town.<br>
- **chas**: Charles River dummy variable (1 if tract bounds river; 0 otherwise).<br>
- **nox**: Nitrogen oxides concentration.<br>
- **rm**: Average number of rooms per dwelling.<br>
- **age**: Proportion of owner-occupied units built before 1940.<br>
- **dis**: Distances to Boston employment centers.<br>
- **rad**: Index of accessibility to radial highways.<br>
- **tax**: Property tax rate per $10,000.<br>
- **ptratio**: Pupil-teacher ratio by town.<br>
- **black**: 1000(Bk - 0.63)^2 where Bk is the proportion of Black residents by town.<br>
- **lstat**: Percentage of lower status population.<br>
- **medv**: Median value of owner-occupied homes in $1000s.<br>
<br>

```{r}
# Load necessary packages and data
# install.packages("MASS")
library(MASS)
data("Boston")

# Display the first few rows of the dataset
head(Boston)
summary(Boston)
```
<br>

## 1.2 Residual Analysis

#### Purpose and Importance
Residual analysis is a crucial step in regression diagnostics. By examining the residuals (the differences between observed and predicted values), we can identify patterns that might indicate model inadequacies, such as non-linearity, heteroscedasticity, or the presence of outliers.<br>
<br>

#### Example: Residual Plots
```{r}
# Fitting a linear model
lm_boston <- lm(medv ~ lstat + rm + age, data = Boston)

# Residual plots
par(mfrow = c(2, 2))
plot(lm_boston)
```
<br>

#### Explanation:
- **Residuals vs. Fitted**: This plot helps detect non-linearity, heteroscedasticity (non-constant variance), and outliers. Ideally, residuals should be randomly scattered around zero.
- **Normal Q-Q**: This plot checks whether the residuals are normally distributed. If the normality assumption holds, the points should be straight.
- **Scale-Location**: Also known as the Spread-Location plot, it checks for homoscedasticity. The residuals should spread equally along the range of fitted values.
- **Residuals vs. Leverage**: This plot helps detect influential observations that might disproportionately affect the model. High leverage points that significantly influence the model will be flagged.
<br>

#### Interpreting Results:
- **Non-random patterns**: In the Residuals vs. Fitted plot, patterns such as a funnel shape suggest heteroscedasticity, while curvature suggests non-linearity.<br>
- **Outliers or Influential Points**: In the Residuals vs. Leverage plot, points with high leverage and large residuals could be influential.<br>
<br>

## 1.3 Checking Assumptions
### Linearity, Normality, and Homoscedasticity

**Linearity**:<br>
The relationship between predictors and the response variable should be linear.<br>
<br>
**Normality**:<br>
The residuals should be normally distributed. The Normal Q-Q plot helps assess this.<br>
<br>
**Homoscedasticity**:<br>
The variance of the residuals should be constant across all levels of the predictors.<br>
<br>

#### Linearity and Homoscedasticity

```{r}
# Checking for linearity and homoscedasticity
plot(lm_boston$fitted.values, lm_boston$residuals)
abline(h = 0, col = "red")
```

::: {.callout-note}
This plot is used to check for linearity and homoscedasticity. Residuals should be randomly scattered around zero without showing any specific pattern.
:::

#### Normality of Residuals
```{r}
# Checking for normality
qqnorm(lm_boston$residuals)
qqline(lm_boston$residuals, col = "red")
```

::: {.callout-note}
Q-Q Plot: If the points lie on the line, the residuals are normally distributed, which satisfies the normality assumption.
:::


## 1.4 Detecting Multicollinearity with VIF

Multicollinearity occurs when the independent variables in a regression model are highly correlated, leading to unreliable estimates of regression coefficients. The Variance Inflation Factor (VIF) detects multicollinearity.

### Calculating VIF

```{r}
# Load the necessary package
# install.packages("car")
library(car)

# Calculate VIF
vif(lm_boston)
```

::: {.callout-note}
**VIF**: A VIF value above 5 or 10 indicates a high level of multicollinearity, suggesting that the variable may need to be removed or combined with others to reduce redundancy.
:::

## 1.5 Testing for Heteroscedasticity

Heteroscedasticity occurs when the variance of residuals is not constant across all levels of the independent variables. This violates the assumption of homoscedasticity, leading to inefficient estimates.

### Breusch-Pagan Test
```{r}
# Load the necessary package
# install.packages("lmtest")
library(lmtest)

# Perform the Breusch-Pagan test
bptest(lm_boston)
```

::: {.callout-note}
**Breusch-Pagan Test**: This test checks for heteroscedasticity in the model. A significant p-value (typically < 0.05) suggests the presence of heteroscedasticity.
:::


## Model Evaluation

Model evaluation assesses how well your regression model fits the data and how well it is likely to perform on new data.

### Metrics for Model Evaluation:
- **R-squared**: Measures the proportion of the variance in the dependent variable that is predictable from the independent variables.
- **Adjusted R-squared**: Adjusts R-squared for the number of predictors, providing a more accurate measure for models with multiple predictors.
- **AIC (Akaike Information Criterion)**: Balances model fit and complexity. Lower AIC values indicate a better model.

#### Comparing Models

```{r}
# Model with additional predictors
lm_boston_full <- lm(medv ~ lstat + rm + age + dis + tax, data = Boston)

# R-squared comparison
summary(lm_boston)$r.squared
summary(lm_boston_full)$r.squared

# Adjusted R-squared comparison
summary(lm_boston)$adj.r.squared
summary(lm_boston_full)$adj.r.squared

# AIC comparison
AIC(lm_boston)
AIC(lm_boston_full)
```

- **R-squared and Adjusted R-squared**: These metrics help assess the explanatory power of your model. Adjusted R-squared is particularly useful when comparing models with different numbers of predictors. <br>
- **AIC**: This criterion helps balance the trade-off between model complexity and goodness of fit. A lower AIC value indicates a model that better fits the data while avoiding overfitting. <br>

::: {.callout-note}
Model Preference: A model with a higher Adjusted R-squared and lower AIC is typically preferred, but it’s also important to consider the predictors' practical significance and the risk of overfitting.
:::


# Session 2: Hands-on Exercises

The diamonds dataset from the ggplot2 package contains data on 53,940 diamonds, with attributes that can be used to explore relationships between price, quality, and physical dimensions.

#### Key Variable Descriptions:
- **carat**: Weight of the diamond (continuous variable).
- **cut**: Quality of the cut, categorized as Fair, Good, Very Good, Premium, and Ideal (ordered factor).
- **color**: Diamond color, with grades from D (best) to J (worst) (ordered factor).
- **clarity**: Clarity of the diamond, ranging from I1 (worst) to IF (best) (ordered factor).
- **depth**: Total depth percentage, which is the height of a diamond divided by its average diameter (continuous variable).
- **table**: Width of the diamond's table expressed as a percentage of its diameter (continuous variable).
- **price**: Price of the diamond in US dollars (continuous variable).

```{r}
# Load the necessary package
library(ggplot2)
data("diamonds")
head(diamonds)
```

Choose a dependent variable (Y) and at least two independent variables (X1, X2, ...). Fit a multiple linear regression model.

## Selecting Variables and Fitting a Regression Model
#### Q1
1) Use the lm() function to create a multiple linear regression model. <br>
2) Display the summary of the model to interpret the coefficients, R-squared, and p-values. <br>

```{r}
# Example (students should choose their own variables):
# Dependent variable: price
# Independent variables: carat, cut, color

model <- lm(price ~ carat + cut + color, data = diamonds)

# Display the model summary
summary(model)
```

## Residual Analysis
#### Q2
1) Create residual plots to check for linearity, homoscedasticity, and normality. <br>
2) Use the par() function to display multiple plots at once. <br>

```{r}
# Residual analysis
par(mfrow = c(2, 2))
plot(model)
```

## Checking for Multicollinearity and Heteroscedasticity
#### Q3
1) Calculate VIF for each independent variable in your model. <br>
2) Use the bptest() function from the lmtest package to test for heteroscedasticity. <br>
3) Interpret the results of the test.

```{r}
# Install the car package if not already installed
# install.packages("car")
library(car)

# Calculate VIF
vif_values <- vif(model)
vif_values
```

```{r}
# Install the lmtest package if not already installed
# install.packages("lmtest")
library(lmtest)

# Breusch-Pagan test for heteroscedasticity
bp_test <- bptest(model)
bp_test
```

## Model Comparison and Evaluation
#### Q4
1) Fit an alternative model with a different set of independent variables. <br>
2) Compare the models based on Adjusted R-squared and AIC to determine which model is better. <br>

```{r}
# Alternative model with different independent variables
model2 <- lm(price ~ carat + clarity + depth, data = diamonds)

# Compare R-squared and Adjusted R-squared
summary(model)$adj.r.squared
summary(model2)$adj.r.squared

# Compare AIC
AIC(model)
AIC(model2)
```

# Session 3: Wrapping Up and Q&A
In this session, we've explored essential concepts in regression diagnostics and model evaluation, including residual analysis, checking assumptions, identifying influential points, and model comparison using R-squared and AIC.<br>
<br>

Practice these techniques with different datasets to better understand and improve your analytical skills. <br>
<br>

**Q&A**: Please share any questions or challenges encountered during the exercises. Let's discuss solutions and clarify concepts!<br>

<br>

### Reference
- https://datageneration.io/dataprogrammingwithr/intro
- Chicago Harris School Coding Camp
- Data.gov, the U.S. government's open data portal
- https://cran.r-project.org/web/packages/MASS/MASS.pdf
- https://ggplot2.tidyverse.org/reference/diamonds.html 

